Poisson approximate likelihood compared to the particle filter

### 2024-09-16 ms.Rnw
some edits
initial dist at t0 should be sum up to pop

init_dist_sbh_mle <- c(
  S1=round(mean(s1_t1_vec),0),
  I1=round(mean(i1_t1_vec),0),
  R1=round(mean(r1_t1_vec),0),
  S2=round(mean(s2_t1_vec),0),
  I2=round(mean(i2_t1_vec),0),
  R2=round(mean(r2_t1_vec),0),
  S3=round(mean(s3_t1_vec),0),
  I3=round(mean(i3_t1_vec),0),
  R3=round(mean(r3_t1_vec),0)
)
@

@@

WWR compared their fitted models with a log-ARMA(2,1) benchmark AIC value of 23043 (line 7).
From this high AIC value, they inferred that all the mechanistic models possessing overdispersion have better statistical fit than a simple loglinear time series model.
However, this AIC value corresponds to SBH's data, not the data fitted by WWR.
Thus, line 7 can properly be compared only to line 1 but not to any other line of  Table~\ref{tab:ovovrealdata}. 
We independently fitted a log-ARMA(2,1) model to the SBH data and obtained a similar AIC value (\Sexpr{myround(sbh_benchmark,tab_signif)}).
Carrying out the same computation for the data fitted by WWR gives a log-ARMA benchmark AIC value of \Sexpr{myround(wwr_benchmark,tab_signif)} (line 8).
Thus, the statistical fit of the mechanistic model considered by WWR is inferior to a simple log-ARMA model.
This holds for all the variants in lines 2--6 of  Table~\ref{tab:ovovrealdata}, regardless of whether PF or PAL is used.
The goal of mechanistic modeling is not necessarily to beat a simple statistical benchmark, but falling far below a simple statistical benchmark is an indication that additional model development could be worthwhile \citep{wheeler24}.
A correct interpretation of Table~\ref{tab:ovovrealdata} is therefore very different to the conclusions drawn by WWR, who compared line 2 inappropriately to lines 1 and 7.

>>>>

sbh_mean_state <- c(
  S1=round(mean(s1_t1_vec),0),
  I1=round(mean(i1_t1_vec),0),
  R1=round(mean(r1_t1_vec),0),
  S2=round(mean(s2_t1_vec),0),
  I2=round(mean(i2_t1_vec),0),
  R2=round(mean(r2_t1_vec),0),
  S3=round(mean(s3_t1_vec),0),
  I3=round(mean(i3_t1_vec),0),
  R3=round(mean(r3_t1_vec),0)
)

#calculate the difference in the init dist
diff <- sum(sbh_mean_state) - pop
#reduce the differenc proportionally
init_dist_sbh_mle <- round(sbh_mean_state - diff*(sbh_mean_state/sum(sbh_mean_state)),0)
#check if init_dist == pop
sum(init_dist_sbh_mle) == pop

#sum(init_dist_sbh_mle) == pop
@

@@

WWR compared their fitted models with a log-ARMA(2,1) benchmark AIC value of 23043 (line 7).
From this high AIC value, they inferred that all the mechanistic models possessing overdispersion have better statistical fit than a simple loglinear time series model.
However, this AIC value corresponds to SBH's data, not the data fitted by WWR.
Thus, line 7 can be properly compared only to line 1 but not to any other line of  Table~\ref{tab:ovovrealdata}. 
We independently fitted a log-ARMA(2,1) model to the SBH data and obtained a similar AIC value (\Sexpr{myround(sbh_benchmark,tab_signif)}).
Carrying out the same computation for the data fitted by WWR gives a log-ARMA benchmark AIC value of \Sexpr{myround(wwr_benchmark,tab_signif)} (line 8).
Thus, the statistical fit of the mechanistic model considered by WWR is inferior to a simple log-ARMA model.
This holds for all the variants in lines 2--6 of  Table~\ref{tab:ovovrealdata}, regardless of whether PF or PAL is used.
The goal of mechanistic modeling is not necessarily to beat a simple statistical benchmark, but falling far below a simple statistical benchmark is an indication that additional model development could be worthwhile \citep{wheeler24}.
A correct interpretation of Table~\ref{tab:ovovrealdata} is therefore very different to the conclusions drawn by WWR, who compared line 2 inappropriately to lines 1 and 7.
